Research Experience

### 2010-04-23 doc/thesis/thesis.tex
more on the lit review

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have adapted
the BOLD model from the initial Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may between
seven \cite{Riera2004} and fifteen \cite{Zheng2002} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

@@

\subsection{BOLD Physiology}
It is will known that Deoxyhemoglobin acts as a contrast agent in EPI imaging
\cite{Buxton1999}, \cite{WEISSKOFF1994}, \cite{Ogawa}, so then what causes
changes Deoxyhemoglobin in a voxel? Obviously increased metabolism will increase
Deoxyhemoglobin, but changes in blood volume in a voxel will also alter 
the MR signal depending on the ratio of Deoxyhemoglobin to hemoglobin. 
This was the impetus
for the ground breaking research \cite{Buxton1999} and \cite{Mandeville1999}.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood flow (CBF) 
and the cerebral blood volume (CBV). In fact \cite{Buxton1999} went to far as
to show that a simple, well chosen blood flow waveform, along with a balloon 
model, could fully account for the BOLD signal. Although complex
changes in metabolism exist (\cite{Zheng2002}), most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Donahue2009}, \cite{Yacoub2006}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the primary signal may be accurately
be estimated by a simple flow-locked expression of the CMRO2. Comparison
between the more complicated models including a separate expression for CMRO2
and the simplified Buxton expression show little actual difference \cite{Deneux2006}.
In \cite{Buxton1999}, it was shown that small changes in the flow expression could
result in vastly different recovery times of the baseline BOLD signal, and
in \cite{Mandeville1999} a cause for delayed CBV recovery was proposed, based
on differences in compliance of capillaries. \cite{Behzadi2005} provided a simple,
complete model and explanation for delayed compliance of neural capillaries,
and those simple changes can result in a significant improvement in the bias error,
as shown in \cite{Deneux2006}. In the end, it may be impossible to refute the
possibility of a prolonged period of increased metabolic activity after activation,
however, much of the BOLD post stimulus undershoot may be explained by a 
prolonged CBV undershoot, as shown in \cite{Chan2009}. Because of the greatly
increased complexity of CMRO2 modeling, compared with the relative simplicity
of the Behzadi model, in most cases it is not worth using the more advanced CMRO2
model unless the metabolic rate is an end goal.

>>>>

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and fifteen \cite{Zheng2002} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

FMRI images as a method of detecting neural activation is based on 
temporal changes of the Blood Oxygen Level Dependent (BOLD) signal.
The BOLD signal is caused by minute changes in the ratio of Deoxygenated
Hemoglobin to Oxygenated Hemoglobin in blood vessels throughout the brain.
Because Deoxygenated hemoglobin is paramagnetic, higher concentrations
attenuate the signal when using T2 weighted imaging techniques, such 
as Echo Planar Imaging (EPI) which is used in FMRI. When axons becomes active,
a large amount of ions quickly flow out of the cell. In order for the action
potential to be used again, an active pumping process moves ions back into the
axon. This process of recharging the axon takes a large amout of enengy, which 
naturally uses oxygen. On a massive scale (cubic millimiter) this activation/recharge
process is happening all
the time; however, it happens at a much higher rate when a portion of
the brain is very active. Thus, blood vessels in a very active area will 
tend to have less oxygenated hemoglobin, and more deoxygenated hemoglobin,
resulting in lower FMRI signal. However, to compensate for activation, muscles that
control blood vessels relax in that region to allow more blow flow, which 
in fact results in a higher concentration of oxygenated hemoglobin. Thus,
increased activation actually tends to \emph{increase} the MR signal in
comparison with the base level. It is this overcompensation that is the 
primary signal detected with FMRI imaging. This cascade of events
can, as a consequence of increased activity, increase the local metabolism, 
blood flow, blood volume, and oxygenated hemoglobin; though not necessarily
in sync. The lag between these
various factors is what causes many of the complexities of the BOLD signal.

@@

\subsection{BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1999}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between DeHb/O2Hb and neural activity is non-trivial. Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is encouraged by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1999}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood flow (CBF) 
and the cerebral blood volume (CBV). In fact \cite{Buxton1999} went to far as
to show that a simple, well chosen blood flow waveform coupled with a simple
cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although complex
models for metabolism exist (\cite{Zheng2002}), most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Donahue2009}, \cite{Yacoub2006}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2. For some time there was 
debate regarding the cause for the post-stimulus undershoot often seen in 
FMRI studies. The problem stemmed from the fact that there were many studies
showing CBV and CBF returning to baseline well before the BOLD signal. In fact
this result seemed to be confirmed by several studes: \cite{Frahm2008}, \cite{Donahue2008}, 
\cite{Lu2004}.

%here

In fact, the original Buxton
paper demonstrated how minimal changes in the flow expression could
result in vastly different recovery times of the baseline BOLD signal, and
in \cite{Mandeville1999} a cause for delayed CBF recovery was proposed, based
on hysterous in the compliance of capillaries. \cite{Behzadi2005} provided
additional incite into this possibility, going so far as proposing a simple
yet complete model and explanation for delayed compliance of neural capillaries.
Those simple changes can result in error levels on par with the more 
complex CMRO2 inclusive models, which was one of the major findings of \cite{Deneux2006}. 
This was reiterated in \cite{Chan2009}, which concluded that many of
the studies claiming a CBV recovery well before the BOLD signal 
(\cite{Frahm1996}, \cite{Kruger1996}, \cite{Lu2004}) 
could actually have been detecting arterial blood flow, which does not significantly effect the
BOLD signal. 
In the end, it may be impossible to refute the
possibility of a prolonged period of increased metabolic activity after activation,
however, much of the BOLD post stimulus undershoot may be explained by a 
prolonged CBV undershoot, as shown in \cite{Chan2009}. Because of the greatly
increased complexity of CMRO2 modeling, compared with the relative simplicity
of the Behzadi model, in most cases it is not worth using the more advanced CMRO2
model unless the metabolic rate is an end goal.
